### 2010-03-17 papers/icfp-2010-desc/paper_desc.tex
Desc@ICFP: general review

In this section, we describe an alternative implementation of
inductive types, as we know them in simply-typed languages such as
Haskell or OCaml. This choice is motivated by pedagogical
considerations: by describing the layman data-types in our setting, we
build on our intuition of data-types we are familiar with. However,
our goal is and remains to discuss data-types as we find them in
dependently-typed languages.

Our proposal is inspired by, but less involved than, Dybjer and
Setzer's closed formulation of
induction-recursion~\cite{dybjer:general-ir, dybjer:axiom-ir,
  dybjer:ir-initial-algebra, dybjer:iir}. In the context of simple
data-types, the full power of induction-recursion is not
necessary. Therefore, we adopt a stripped-down version of it. As we
extend the scope of our universe of data-types, our presentation will
diverge from induction-recursion. We shall compare both approaches in
Section~\ref{sec:discussion}

@@

In dependently-typed language, $\Sigma$-types can be read in two
rather different ways. This duality is actually reflected in the
notation we can find in the literature, depending on the sensibility
of the author. Hence, the type $\SigmaTy{A}{B}$ is sometimes written
$\Sigma_{x : A} (B\: x)$. This notation stresses the fact that
$\Sigma$-types are a generalisation of sums over arbitrary
arities. When simply-typed languages have finite sums,
dependently-typed languages have sums indexed by any set.

@@

In the simply-typed world, the essence of data-type definition can be
summed up by the term ``sum-of-product''. A data-type is defined by a
finite sum -- a choice -- of constructors, each of them is composed by
a product -- a tuple -- of arguments. Therefore, the grammar of
data-types definition in simply-typed languages, such as Haskell or
OCaml, follow this form.

To model these data-types, we are simply left with capturing this
grammar in a dependently-typed setting. Hence, the notion of
``sum-of-product'' naturally translates into
\emph{sigmas-of-sigmas}. However, beyond this intuition, it should be
clear that a direct encoding of data-types through raw $\Sigma$-types
is not a viable option. Indeed, just as for finite sets, encoding
throw away information when we crucially need it. In the realm of
data-types, things become even more tougher, as one could legitimately
expect an induction principle, for instance. Anonymous in a see of
$\Sigma$s, our data-types have little chance of survival.

@@

For our sigma-of-sigmas to be known, we crucially need the ability to
name them. To this end, we will use a standard technique in
dependently-typed programming: we construct an ad-hoc universe. This
technique dates back to Martin-L\"of definition of type
theory~\cite{martin-lof:itt}. Since then, it has been fruitful as a
programming technique~\cite{benke:universe-generic-prog, whoelse?}
\note{I would need some other example of universe construction for
  practical purposes}. We refer the reader to Agda's
tutorial~\cite{norell:agda-tutorial} for a pedagogical presentation of
universe construction.

The key idea behind universe construction is our ability to make names
by defining new types. These names are called \emph{codes}. By
defining a set of codes, we somehow define the syntax of a
language. However, as such, a system of code is useless as it lacks a
semantics. Instead of equipping the universe of codes with some
computational behaviour, we pragmatically chose to \emph{interpret}
these codes back into the standard type theory. Hence, codes act only
as labels, while the type theory provides the computational machinery.

Codes being simple labels, we have the ability to inspect them, hence
taking advantage of their structural information. Being able to
inspect them, we are therefore free to compute over them: deriving new
codes from previous ones, or even new functions on them. In several
occasions, we will have the opportunity to witness the power of
universes.

\note{Is that really a motivating motivation for the usage of universe
  construction? }

@@

Hence, we propose to embed inductive types as a universe in our
dependent type theory, the universe of \emph{descriptions}
(Figure~\ref{fig:desc_universe}). As expected, the code of this
universe mimics the standard grammar of data-types definitions in
simply-typed languages. Hence, we have a $\DSigma{}{}$ code,
interpreted as a $\Sigma$-type, to build the
sigmas-of-sigmas. Descriptions are terminated by $\DUnit$, which
contains no useful payload. The functoriality of the data-types is
introduced by $\DIndx$. When we tie the knot with a fix-point, the
hole left open by $\DIndx$ will be turned into a standard recursive
argument. We notice that this functoriality appears in the type of
$\descop{D}{}$ itself, for a given data-type definition $D$. This
corresponds to the object part of the functor described by $D$.

@@

Let us explain its construction. First, we use $\DSigma{}{}$ to build
a sum between $\NatZero$ and $\NatSuc{}$. In the $\NatZero$ case, we
reach the end of the description: there is no useful payload in that
case. In the $\NatSuc{}$ case, we left one hole open for the recursive
argument, and we close the description.

In a more synthetic notation, we have simply implemented the following
functor:

@@

The modification consists in turning the $\NatSuc{}$ constructor into
a proper $\ListCons$ taking an argument in X followed by a recursive
argument. In this case, we use $\DSigma{}{}$ in its product
interpretation: we pack an element of $X$ together with the recursive
argument. Easily, one sees that this code actually defines the
following functor:

@@

From the examples above, we observe that data-types are defined by a
$\DSigma$ which first argument is a finite set of constructors. The
descriptions fitting into this pattern are called \emph{tagged}
description. Formally, we have:

\note{We have talked about a tagged description based on $\List (\UId
  \times \Desc)$. It makes the free monad easier to construct but,
  imho, considerably clutters anything else. I've reverted back to a
  more sigma-friendly presentation.}

@@

Again, this is a clear reminiscence of the ``sum-of-products'' style:
we have a finite sum of constructors. Of course, every description can
be expressed into this style, using a singleton as single constructor
if necessary.

\note{I think we need a function mapping a $\TagDesc$ to the
  corresponding $\Desc$. This function should go here.}

@@

So far, we have used our universe of descriptions to build
functors. To illustrate its usage, we have implemented the signature
functors of natural numbers, lists, and binary trees. However, the
class of functors expressible with descriptions enjoy another
property: they all admit a fix-point.

\note{ I'm afraid that this section is killed by my categorical gibberish. }

Let us constructively prove this. First of all, for an object $D$
describing a signature functor, we define its fix-point by the type
former $\Mu{D}$. Inhabitants of the fix-point are materialised by the
structure map: $\Con{x}$ inhabits $\Mu{D}$ if $x$ belongs to the
interpretation of $D$ with the carrier of the initial algebra:

@@

For the less categorically inclined reader, we can simply read this
construction as \emph{tying the knot}: the holes left open by the
functor are filled by its own recursive definition. Hence, we have
recovered the data-types we are usually able to define in simply-typed
languages.

@@

Together with the type formers defined above, we would legibly expect
an elimination principle. Following the categorical intuition, we
could be tempted to provide it as a catamorphism:

@@

Whereas this definition is amply sufficient in the simply-typed world,
it comes short of its promises in the dependently-typed realm. Indeed,
this type is absolutely non dependent. As this operator will be the
corner stone of any inductive definition, we are better off sharping
it before use.

@@

At this stage, we are still left with implementing this
$\F{induction}$ principle. For space reason, we refer the reader to
our technical report for the actual code, hence the proof of (weak)
initiality. Having this dependent catamorphism, we are empowered with
the ability to $\emph{compute}$ over descriptions. Note that this
induction principle is the first manifestation of a generic operation
over descriptions. Thanks to $\F{induction}$, any data-type we define
automatically comes equipped with an induction principle.

@@

In this section, we have introduced a universe of description. We have
shown how to manipulate these codes to build data-types. We have also
presented a fix-point construction and its associated dependent
catamorphism. However, one could argue that programming with these
objects is not practical.

@@

This concludes our presentation of the universe of descriptions. In
this section, we have seen how data-types from the simply-typed world
can be expressed in our dependent-type setting. By relying on a
universe construction technique, we benefit from the ability to
inspect and manipulate these codes. This opens some appealing
opportunities for generic programming. Finally, we have demonstrated
that programming with codes is practical. To this end, we have
extended the type propagation machinery. In this setting, we can
define, interact, and build data-types as transparently as in the
simply-typed world.

>>>>

In this section, we describe an implementation of inductive types, as
we know them in simply-typed languages such as Haskell or OCaml. This
choice is motivated by pedagogical considerations: by describing the
layman data-types in our setting, we build on our intuition of
data-types we are familiar with. However, our goal is and remains to
discuss data-types as we find them in dependently-typed languages.

Our proposal is inspired by, but less involved than, Dybjer and
Setzer's closed formulation of
induction-recursion~\cite{dybjer:general-ir, dybjer:axiom-ir,
  dybjer:ir-initial-algebra, dybjer:iir}. In the context of simple
data-types, the full power of induction-recursion is not
necessary. Therefore, we adopt a stripped-down version of it. As we
extend the scope of our universe of data-types, our presentation will
diverge from induction-recursion. We shall compare both approaches in
Section~\ref{sec:discussion}.

@@

In dependently-typed language, $\Sigma$-types can be interpreted in
two different ways. This duality is actually reflected in the notation
we can find in the literature, depending on the sensibility of the
author. Hence, the type $\SigmaTy{A}{B}$ is sometimes written
$\Sigma_{x : A} (B\: x)$. This notation stresses the fact that
$\Sigma$-types are a generalisation of sums over arbitrary
arities. When simply-typed languages have finite sums,
dependently-typed languages have sums indexed by any set.

@@

In the simply-typed world, the essence of data-type definition can be
summed up by the expression ``sum-of-product''. A data-type is defined
by a finite sum -- a choice -- of constructors, each of them is
composed by a product -- a tuple -- of arguments. This form governs
the grammar of data-type definition in languages such as Haskell or
OCaml.

To model these data-types, we simply need to capture this grammar in
our dependently-typed setting. Here, the notion of ``sum-of-product''
naturally translates into \emph{sigmas-of-sigmas}. However, beyond
this intuition, it should be clear that a direct encoding of
data-types through raw $\Sigma$-types is not a viable option. Indeed,
just as for finite sets, encoding throw away information when we
crucially need it. In the realm of data-types, things become even more
tougher, as one could legitimately expect an induction principle, for
instance. Anonymous in a see of $\Sigma$s, our data-types have little
chance of survival. \note{Ridiculous tentative of poetry. I was
  really tired when I wrote that.}

@@

For our sigmas-of-sigmas to be known, we crucially need the ability to
name them. \note{Pursuing this crappy analogy. Solving the type
  presentation/representation debate would help solving that point.}
To this end, we will use a standard technique in dependently-typed
programming: we construct an ad-hoc universe. This technique dates
back to Martin-L\"of definition of type
theory~\cite{martin-lof:itt}. Since then, it has been fruitful as a
programming technique~\cite{benke:universe-generic-prog, whoelse?}
\note{I would need some other example of universe construction for
  practical purposes}. We refer the reader to the Agda
tutorial~\cite{norell:agda-tutorial} for a pedagogical presentation of
universe construction. \note{Is that really a motivating motivation
  for the usage of universe construction? }

The key idea behind universe construction is our ability to make names
by defining new types. These names are called \emph{codes}. By
defining a set of codes, we somehow define the syntax of a
language. However, as such, a system of code is useless as it lacks a
semantics. Instead of equipping the universe of codes with some
computational behaviour, we pragmatically chose to \emph{interpret}
these codes back into the standard type theory. Hence, codes act only
as labels, while the type theory provides the computational
machinery. \note{Is that a good description?}

Codes being simple labels, we have the ability to inspect them, hence
taking advantage of structural information. Being able to inspect
them, we are therefore free to compute over them: deriving new codes
from previous ones, or even new functions on them. In several
occasions, we will have the opportunity to witness the power of
universes.

@@

We propose to embed inductive types as a universe in our dependent
type theory, the universe of \emph{descriptions}
(Figure~\ref{fig:desc_universe}). As expected, the code of this
universe mimics the standard grammar of data-types definitions in
simply-typed languages. Hence, we have a $\DSigma{}{}$ code,
interpreted as a $\Sigma$-type, to build the
sigmas-of-sigmas. Descriptions are terminated by $\DUnit$, which
contains no useful payload. The functoriality of the data-types is
introduced by $\DIndx$. When we tie the knot with a fix-point, the
hole left open by $\DIndx$ will be turned into a standard recursive
argument. We notice that this functoriality appears in the type of
$\descop{D}{}$ itself, for a given data-type definition $D$. This
corresponds to the object part of the functor described by $D$.

@@

Let us explain its construction. First, we use $\DSigma{}{}$ to give a
choice between the $\NatZero$ and $\NatSuc{}$ constructors. In the
$\NatZero$ case, we reach the end of the description: there is no
useful payload in that case. In the $\NatSuc{}$ case, we left one hole
open for the recursive argument, and we close the description.

In a more synthetic notation, we have described the following functor:

@@

The modification consists in turning the $\NatSuc{}$ constructor into
a proper $\ListCons$ taking an argument in X followed by a recursive
argument. In this case, we use $\DSigma{}{}$ in its product
interpretation: we pack an element of $X$ together with the recursive
argument. Easily, one sees that this code actually describes the
following functor:

@@

From the examples above, we observe that data-types are defined by a
$\DSigma$ which first argument is a finite set of constructors. The
descriptions fitting into this pattern are called \emph{tagged}
description. Formally, we have:
\note{We have talked about a tagged description based on $\List (\UId
  \times \Desc)$. It makes the free monad easier to construct but,
  imho, considerably clutters anything else. I've reverted back to a
  more sigma-friendly presentation.}

@@

Again, this is a clear reminiscence of the ``sum-of-products'' style:
we have a finite sum of constructors. Of course, every description can
be expressed in this style, using a singleton as single constructor if
necessary.
\note{I think we need a function mapping a $\TagDesc$ to the
  corresponding $\Desc$. This function should go here.}

@@

So far, we have used our universe of descriptions to build
functors. To illustrate its usage, we have implemented the signature
functors of natural numbers, lists, and binary trees. However, the
class of functors expressible with descriptions enjoy another
property: they all admit a fix-point. Indeed, descriptions are
interpreted into \emph{polynomial functors}~\cite{who?}\note{who is
  the relevant source for polynomial functors?}, as they are only
composed of $\Sigma$-types, the unit type, and variables.

We construct the fix-point by \emph{tying the knot} of the
interpretation: the holes left open by the functor are filled by its
own recursive definition:

@@

At this stage, we have recovered the data-types we are usually able to
define in simply-typed languages.

@@

Together with the type formers defined above, we would legibly expect
an elimination principle. Following a categorical intuition, we could
be tempted to provide it as a catamorphism:

@@

Whereas this definition is amply sufficient in the simply-typed world,
it comes short of its promises in the dependently-typed realm. Indeed,
this type is not dependent. As this operator will be the corner stone
of any inductive definition, we are better off sharping it before use.

@@

At this stage, we are still left with implementing this
$\F{induction}$ principle. For space reason, we refer the reader to
our technical report for the actual code. Having this dependent
catamorphism, we are empowered with the ability to $\emph{compute}$
over descriptions. Note that this induction principle is the first
manifestation of a generic operation over descriptions. Thanks to
$\F{induction}$, any data-type we define automatically comes equipped
with an induction principle.

\subsection{Extending type propagation}

@@

So far, we have introduced a universe of description. We have shown
how to manipulate these codes to build data-types. We have also
presented a fix-point construction and its associated dependent
catamorphism. However, one could argue that programming with these
objects is not practical.

@@

This concludes our presentation of the universe of descriptions. In
this section, we have seen how data-types from the simply-typed world
can be expressed in our dependently-typed setting. By relying on a
universe construction technique, we benefit from the ability to
inspect and manipulate these codes. This opens some appealing
opportunities for generic programming. Finally, we have demonstrated
that programming with codes is practical. To this end, we have
extended the type propagation machinery. In this setting, we can
define, interact, and build data-types as transparently as in the
simply-typed world.

### 2010-03-17 papers/icfp-2010-desc/paper_desc_levitation.tex
Desc@ICFP: general review

In this section, we will fulfil several promises. One promise we made
was to effectively implement the code of the $\Desc$ universe. Another
promise we tacitly made is to implement the type former of finite
sets. Despite being a perilous pedagogical exercise for the authors,
we hope to convey to the reader the dizzy feeling of levitation,
without the falling.

@@

In Section~\ref{sec:finite-sets}, we have promised an implementation
of finite sets. It consists in the objects specified in
Figure~\ref{fig:typing-finite-set}. In particular, we will now
implement the $\EnumU$ type former and its constructors. Let us recall
their specification:

@@

The names $\NilE$ and $\ConsE{}{}$ are absolutely not a
coincidence. $\EnumU$ is indeed simply isomorphic to a list, with
$\UId$s as parameter. However, in Section~\ref{sec:desc-examples}, we
have shown how to build lists in the universe of
description. Therefore, the actual implementation of $\EnumU$ sums up
to the following, trivial definition:

@@

Apart from these changes, we are left with implementing the other
components of finite sets. This consists in $\EnumT$, $\Ze$, $\Su$, as
well as the $\F{switch}$ operator. Note that the actual implementation
of $\EnumU$ does not influence our implementation of $\EnumT$: be they
hard-coded or codes in the $\Desc$ universe, the $\EnumU$ objects just
behave similarly. We have witnessed this effect when carrying this
operation in Epigram, moving from an hard-coded presentation to a
self-hosted one. Absolutely no changes to the $\EnumT$ world was
required.

@@

In this section, we have replaced a low-level presentation of finite
sets by a self-hosted one, expressed in the universe of
descriptions. However, formally, the content of the type theory
remains unchanged: objects which were present before the modification
are still there. Conversely, we have not introduced any spurious
objects in the type theory.

@@

We shall now fulfil our promises. We are going to implement the codes
of the universe of descriptions. Interestingly, in and by itself, a
code is nothing but a data-type. We are in the same situation than
with $\EnumU$: we ought to be able to describe the codes of $\Desc$ in
$\Desc$ itself. Hence, this code would be a first-class citizen, born
with the standard, generic equipment of data-types.

@@

Let us explain how we have proceeded and the obstacle we face. Much as
the data-types we have seen so far, we first have to choose a
constructor. Constructors correspond to the code's cases: $\DUnit$,
$\DSigma$, and $\DIndx$. The reader will notice that the tagged
notation we have used so far for $\Desc$ constructors now fully make
sense: these were actually the tags we are defining. \note{This
  sentence badly needs a rephrasing.}

The case of $\DUnit$ is trivial: we immediately reach the end of the
data-type. $\DUnit$ has no payload. The case of $\DIndx{}$ comes
naturally: it only takes a recursive argument -- a description. The
case of $\DSigma$ is problematic. Recall the definition of $\DSigma$:

@@

In a first glance, we have achieved our goal. We have described the
codes of the universe of description. Taking the fix-point of this
object gives us $\Desc$, up to isomorphism. We have implemented our
code of data-types as a levitating object inside itself. However, this
levitation operation, just as any magic trick, relies on an invisible
cable. Let us reveal it, hence finishing our implementation.

@@

Our reader might be slightly confused to learn that the trick is
visible in the definition of $\DescD$
(Fig.\ref{fig:desc-levitate}). Or rather, it is made invisible by
careful usage of type propagation. Indeed, let us elaborate this term
down to the low-level type theory. The $\EnumT [ \ldots ]$ construct
elaborates to a finite set $e$ in $\EnumU$, inhabited by the
codes. Then, we can type-check the case definition, between square
brackets $[ \ldots ]$. This term $f$ is pushed into the type
$\EnumT{e} \To \Desc$. This corresponds to a finite function
definition we have implemented in Figure~\ref{fig:type-checking}.

@@

\[
\PLAM{x}{(\EnumT{e})} \switch{e}{\LAM{\_} \Mu{\DescD}}{\pi^f}{x}
\]

But this is rather problematic! We are still in the process of
constructing $\DescD$, and the motive of $\F{switch}$ is abruptly
begging for an $\eta$-expanded $\Desc$. Despite our willingness, we
will never be able to materialise such motive.

@@

The entire work of the magician stands here, in this extension. One
could be worried by the availability of such operations in our type
theory. Indeed, it would be inconvenient for the user to have to
decide between two variants. The solution to this problem stands, once
again, in the type propagation system. Just as the $\F{switch}$
operator, $\F{switchD}$ can be automatically elaborated. To this end,
we simply extend the type-checking rules
(Fig.~\ref{fig:type-checking}) with the following inference rule:

\[
\Rule{\Gamma \Vdash \propag{\push{t}{\spiD{e}}}
                           {t'}}
     {\Gamma \Vdash \begin{array}{@{}l} 
                        \propag{\push{t}{\EnumT{e} \To \Desc}}
                               {\\ \PLAM{x}{(\EnumT{e})} \switchD{e}{t'}{x}}
                    \end{array}}\;\mbox{t is $[]$ or $[a,b]$}
\]

@@

This concludes our levitation work. Beyond its pedagogical value, this
exercise has several practical outcomes. First of all, this exercise
reveals that the $\Desc$ universe is just plain data. Just as any
piece of data, it can therefore be inspected and manipulated. However,
unlike any data, it is expressed in the $\Desc$ universe. As a
consequence, it is equipped, for free, with an induction
principle. So, our ability to inspect and program with $\Desc$ is not
restricted to the meta-language: we now have all the necessary
equipment in the target language to \emph{program} over data-types. In
this setting, generic programming is just about \emph{programming}.

@@

The catamorphism can be characterised as an induction on the
description $D$, with a non-dependent motive targeting $T$. Given a
node $xs$ and the induction hypotheses, the method ought to build an
element of $T$. Provided we know how to make an element of
$\descop{D}{T}$, this step will be performed by the algebra $f$. Let
us take a look at this jigsaw:

@@

With this example, we have shown how we can derive a generic
operation, the catamorphism, from a pre-existing generic operation,
the induction principle. This has been made possible by our ability to
manipulate descriptions as first-class objects: the catamorphism is,
basically, a function mapping a $\Desc$ to a data-type specific
operation. Having the ability to program on $\Desc$ is the key to
define such function. The induction principle of $\Desc$ -- inherited
by levitation -- plays a central role here: without an induction
principle, it would be impossible to program on $\Desc$.

@@

\newcommand{\FMFreeMonad}{\D{FreeMonad}}
\newcommand{\FMFreeMonadF}{\D{FreeMonad$_{\CN{Free}}$}}
\newcommand{\FMVar}{\C{Var}}
\newcommand{\FMComposite}{\C{Composite}}

@@

\[
\FMFreeMonadF\: F\: X\: Z \mapsto X + F Z
\]

@@

In our setting, the functor will be represented by a tagged
description. The free monad construction will take such tagged
description, a set $X$ of variables, and will compute the tagged
description of the corresponding free monad.

Implementing this function is surprisingly easy:

@@

We simply add a constructor, $\DVar{}$, and define its argument to be
a $\DSigma{X}{\DUnit}$, that is an element of $X$. We keep $E$ and $D$
as they were, hence leaving the functor unchanged. The interpretation
of this definition convince us that we have built the functor
signature of the free monad. The fix-point operation ties the knot and
gives us the full-blown free monad construction.

@@

Our task is therefore to implement the algebra of the
catamorphism. Intuitively, its role is to catch appearance of
$\DVar{x}$, to replace it by $\sigma x$. Otherwise, it should simply
appeal to the induction hypothesis. This corresponds to the following
definition:

@@

Filling the $???$ sub-goal with $\apply\: D\: X\: Y\: \sigma$ finishes
the implementation. To sum up, we have implemented the free monad
construction for an arbitrary tagged description. This gives our user
the ability, for any data-type, to extend it with a notion of
variable. Then, we have equipped this structure with the
corresponding monadic operation, ``bind'' and ``return''. Whereas the
``return'' is a trivial variable former, the ``bind'' operation
corresponds to the substitution of variables by terms. We have shown
how to implement it, for any free monad.

>>>>

In this section, we will fulfil several promises. One promise we made
was to effectively implement the code of the $\Desc$ universe. Another
promise we tacitly was the existence of the type formers for finite
sets. Despite being a perilous pedagogical exercise for the authors,
we hope to convey to the reader the dizzy feeling of levitation,
without the falling.

@@

In Section~\ref{sec:finite-sets}, we have given a specification of
finite sets. It consists in the objects presented in
Figure~\ref{fig:typing-finite-set}. We are going to implement the
$\EnumU$ type former and its constructors. Let us recall their
specification:

@@

The names $\NilE$ and $\ConsE{}{}$ are absolutely not a
coincidence. $\EnumU$ is indeed isomorphic to a list of $\UId$s. In
Section~\ref{sec:desc-examples}, we have shown how to build lists in
the universe of description. Therefore, the actual implementation of
$\EnumU$ boils down to the following, trivial definition:

@@

Apart from these changes, we are left with implementing the other
components of finite sets. This consists in $\EnumT$, $\Ze$, $\Su$, as
well as the $\F{switch}$ operator. Note that the actual implementation
of $\EnumU$ does not influence our implementation of $\EnumT$: be they
hard-coded or codes in the $\Desc$ universe, the $\EnumU$ objects just
behave similarly. We have witnessed this effect when carrying this
operation in Epigram, moving from an hard-coded presentation to a
self-hosted one. Absolutely no change to the $\EnumT$ objects was
required.

@@

In this section, we have replaced a low-level presentation of finite
sets by a self-hosted one, expressed in the universe of
descriptions. However, formally, the content of the type theory
remains unchanged: objects which were present before the modification
are still there. Conversely, we have not introduced any spurious
object.

@@

We shall now fulfil our promises for $\Desc$. We are going to
implement the codes of the universe of descriptions. Interestingly, in
and by itself, a code is nothing but a data-type. We are in the same
situation than with $\EnumU$: we ought to be able to describe the
codes of $\Desc$ in $\Desc$ itself. Hence, this code would be a
first-class citizen, born with the standard, generic equipment of
data-types.

@@

Let us explain how we have proceeded and the obstacle we face. Much as
the data-types we have seen so far, we first have to choose a
constructor. Constructors correspond to the code's constructors:
$\DUnit$, $\DSigma$, and $\DIndx$. The reader will notice that the
tagged notation we have used so far for $\Desc$ constructors now fully
make sense: these were actually the tags we are defining. \note{This
  sentence badly needs a rephrasing.}

The case of $\DUnit$ is trivial: we immediately reach the end of the
data-type. $\DUnit$ has no payload. The case of $\DIndx{}$ comes
naturally: it only takes a recursive argument -- a description. The
case of $\DSigma$ is problematic. Recall the specification of
$\DSigma$:

@@

In a first glance, we have achieved our goal. We have described the
codes of the universe of description. Taking the fix-point of this
object gives us $\Desc$, up to isomorphism. We have implemented our
code of data-types as an object levitating inside itself. However,
this levitation operation, just as any magic trick, relies on an
invisible cable. Let us reveal it, hence finishing our implementation.

@@

Our reader might be slightly confused to learn that the trick is
visible in the definition of $\DescD$
(Fig.\ref{fig:desc-levitate}). Or rather, it is made invisible by
careful usage of type propagation. Indeed, let us try to elaborate
this term down to the low-level type theory. The $\EnumT [ \ldots ]$
construct elaborates to a finite set $e$ in $\EnumU$, inhabited by the
codes. Then, we can type-check the case definition, between square
brackets $[ \ldots ]$. This term $f$ is pushed into the type
$\EnumT{e} \To \Desc$. This corresponds to a finite function
definition, as formalised in Figure~\ref{fig:type-checking}.

@@

\[
\PLAM{x}{(\EnumT{e})} \switch{e}{(\LAM{\_} \Mu{\DescD})}{\pi^f}{x}
\]

But this is quite problematic. We are still in the process of
constructing $\DescD$, and the motive of $\F{switch}$ is abruptly
begging for this very same $\DescD$. Despite our willingness, we
cannot materialise such motive.

@@

The entire work of the magician stands here, in this extension. One
could be worried by the availability of such operations in our type
theory. Indeed, it would be inconvenient for the user to have to
decide between two variants. The solution to this problem is, once
again, to use the type propagation system. Just as the $\F{switch}$
operator, $\F{switchD}$ can be automatically elaborated. To this end,
we extend the type-checking rules (Fig.~\ref{fig:type-checking}) with
the following inference rule:

\[
\Rule{\Gamma \Vdash \propag{\push{t}{\spiD{e}}}
                           {t'}}
     {\Gamma \Vdash \begin{array}{@{}l}
                        \propag{\push{t}{\EnumT{e} \To \Desc}}
                               {\\ \PLAM{x}{(\EnumT{e})} \switchD{e}{t'}{x}}
                    \end{array}
     }\;\mbox{t is $[]$ or $[a,b]$}
\]

@@

This concludes our levitation work. Beyond its pedagogical value, this
exercise has several practical outcomes. First of all, this exercise
reveals that the $\Desc$ universe is just plain data. Just as any
piece of data, it can therefore be inspected and
manipulated. Moreover, it is expressed in the $\Desc$ universe. As a
consequence, it is equipped, for free, with an induction
principle. So, our ability to inspect and program with $\Desc$ is not
restricted to the meta-language: we now have all the necessary
equipment in the target language to \emph{program} over data-types. In
this setting, \emph{generic programming is just about programming}.

@@

The catamorphism can be characterised as an induction on the
description $D$, with a non-dependent motive targeting $T$. Given a
node $xs$ and the induction hypotheses, the method ought to build an
element of $T$. Provided that we know how to make an element of
$\descop{D}{T}$, this step will be performed by the algebra $f$. Let
us take a look at this jigsaw:

@@

With this example, we have shown how we can derive a generic
operation, the catamorphism, from a pre-existing generic operation,
the induction principle. This has been made possible by our ability to
manipulate descriptions as first-class objects: the catamorphism is,
basically, a function mapping a $\Desc$ to a data-type specific
operation.

@@

\newcommand{\FMFreeMonad}{\D{FreeMonad}}
\newcommand{\FMFreeMonadD}{\D{FreeMonadD}}
\newcommand{\FMVar}{\C{Var}}
\newcommand{\FMComposite}{\C{Composite}}

@@

\[
\FMFreeMonadD\: F\: X\: Z \mapsto X + F Z
\]

@@

In our setting, the functor will be represented by a tagged
description. The free monad construction will take such tagged
description, a set $X$ of variables, and will compute the tagged
description of the corresponding free monad. Implementing this
function is surprisingly easy:

@@

We simply add a constructor, $\DVar{}$, and define its argument to be
a $\DSigma{X}{\DUnit}$, that is an element of $X$. We keep $E$ and $D$
as they were, hence leaving the functor unchanged. Unfolding the
interpretation of this definition, we convince ourself that this
corresponds to the functor $\FMFreeMonadD$. The fix-point operation
ties the knot and gives us the full-blown free monad construction.

@@

Our task is therefore to implement the algebra of the
catamorphism. Intuitively, its role is to catch appearances of
$\DVar{x}$ and replace them by $\sigma x$. Otherwise, it should simply
appeal to the induction hypothesis. This corresponds to the following
definition:

@@

Filling the $???$ sub-goal with $\apply\: D\: X\: Y\: \sigma$
completes the implementation. To sum up, we have implemented the free
monad construction for an arbitrary tagged description. This gives our
user the ability, for any data-type, to extend it with a notion of
variable. Then, we have equipped this structure with the corresponding
monadic operation, ``bind'' and ``return''. Whereas the ``return'' is
a trivial variable former, the ``bind'' operation corresponds to the
substitution of variables by terms. We have shown how to implement it,
for any free monad.
